---
title: "DSCI 610: Analysis of Time-to-Event Data"
output: html_notebook
---

## Install needed packages and set working directory:
```{r}
library("tidyverse")
library("survival")
library("ggfortify")
library("tidyr")

setwd("C:\\Users\\cy_su\\DataspellProjects\\DSCI_610_Data_Analytics_for_Health_Sciences\\Week_13")
```

## Import the data:
```{r}
ovarian_cancer_txt <- read_lines("oc.txt", skip = 15)
ovarian_cancer_df <- data.frame(ovarian_cancer_txt)
ovarian_cancer_df <- separate(ovarian_cancer_df, col = ovarian_cancer_txt, into = c('Age', 'Grade', 'Stage', 'Residual_Disease', 'Treatment', 'Time_to_Reoccurance', 'Event'))

```

## 1.
    Consider the data in the data set (oc.txt) on time (in days) from the second look laparotomy
    (2LL) to recurrence of ovarian cancer. Please read the header of the data file for a description of the
    variables.

    (a) Plot the Kaplan-Meier estimates of the survival curves for each of the two treatment groups on
    one plot. Calculate 95% confidence intervals for the survival functions at 365 days and 1096 days
    (i.e., 1 and 3 yr survival).

    (b) Perform the logrank test comparing the survival experience of the two groups. Interpret the results.
```{r}
time <- as.integer(ovarian_cancer_df[, 6]) #time in days
delta <- as.integer(ovarian_cancer_df[, 7])
group <- as.integer(ovarian_cancer_df[, 5])

fit <- survfit(Surv(time, delta) ~ group, conf.type = "log-log")

summary(fit)
```

### Plot the survival estimates
```{r}
autoplot(fit, surv.linetype = 'dashed', censor.shape = '*', conf.int = FALSE, censor.size = 5)
```

## 2.
    Following surgical treatment of ovarian cancer, patients may undergo a course of chemotherapy.
    In a study of two different forms of chemotherapy treatment (Edmunson, 1979), researchers compared the anti-tumor
    effects of cyclophosphamide alone and cyclophosphamide combined with adriamycin.

    The trial involved 26 women with minimal residual disease and who had experienced surgical excision
    of all tumor masses greater than 2 cm in diameter. Following the surgery, the patients were further
    classified according to whether the residual disease was completely or partially excised.
    The age of the patient and their performance status were also recorded at the beginning of the trial.
    The response variable was the survival time in days following randomization to one or other of the two
    chemotherapy treatments. The variables in the data set (Chemotherapy in ovarian cancer patients.dat)
    are as follows:
    Time: Survival time in days
    Status: Event indicator (0 = censored, 1 = uncensored)
    Treat: Treatment (1=single, 2= combined)
    Age: Age of patient in years
    Rdisease: Extent of residual disease (1= incomplete, 2 = complete)
    Perf: Performance status: (1 = good, 2 = poor)

    Fit a Cox proportional hazards model adjusting for age, extent of residual disease, and performance.
    Interpret the results.

    Read in the data and clean the data:
```{r}
chemotherapy_data <- read.delim('Chemotherapy in ovarian cancer patients.dat', header = TRUE, sep = "")
chemotherapy_data
mean(chemotherapy_data$age)
chemotherapy_data_age_mutated <- mutate(chemotherapy_data, age.55 = age - 55)
chemotherapy_data_age_mutated
```

### Cox Proportional Hazards Model Build for Regular Age:
```{r}
coxph.M_1 <- coxph(Surv(time, status) ~ age + rdisease + perf, data = chemotherapy_data)
summary(coxph.M_1)
```

### Cox Proportional Hazards Model Build for 55-year-old:
```{r}
coxph.M_2 <- coxph(Surv(time, status) ~ age.55 + rdisease + perf, data = chemotherapy_data_age_mutated)
summary(coxph.M_2)
```

### Interpretation
    1. β1 = 0.142. This means that the difference in the log hazard of a patient being 56 and a patient being 55 is 0.142. This means that being 56 is more of a hazard then being 55
    2. β2 = 0.674. This means that the difference in the log hazard of a patient having a complete residual disease and a pateint having an incomplete residual disease is 0.674. This means that having a complete residual disease is more hazardous than having a not complete residual disease.
    3. β3 = 0.215. This means that the difference in the log hazard of a patient having poor performence and a patient having good performence is 0.215. This means that having poor performence is more hazardous then having good performence.

### Likelihood Ratio Test
```{r}
coxph.M_3 <- coxph(Surv(time, status) ~ rdisease + perf, data = chemotherapy_data_age_mutated)
anova(coxph.M_3, coxph.M_2)
```
```{r}
coxph.M_4 <- coxph(Surv(time, status) ~ age.55 + perf, data = chemotherapy_data_age_mutated)
anova(coxph.M_4, coxph.M_2)
```
```{r}
coxph.M_5 <- coxph(Surv(time, status) ~ age.55, data = chemotherapy_data_age_mutated)
anova(coxph.M_5, coxph.M_2)
```
```{r}
summary(coxph.M_5)
```
### Interpretation of final model
    Since the liklihood of the ratio test with extent of residual disease and perofrmence were not significant, they have been removed from the final model leaving only age.

    1. β1 = 0.161. This means that the difference in the log hazard of a patient being 56 and a patient being 55 is 0.161. This means that being 56 is more of a hazard then being 55