---
title: "DSCI_610_HW_4_R_Code_Cody_York"
output: html_notebook
---

Import needed libraries:
```{r}
library(tidyverse)
library(kableExtra)
```

Import the datasets:

I am skipping the first line since it is just something that came with the csv.
```{r}

nui_2014 <- read.csv('Data/nui_pt_all_counties_IN_2014.csv', header = TRUE, skip = 1)
nui_2015 <- read.csv('Data/nui_pt_all_counties_IN_2015.csv', header = TRUE, skip = 1)
nui_2016 <- read.csv('Data/nui_pt_all_counties_IN_2016.csv', header = TRUE, skip = 1)
nui_2017 <- read.csv('Data/nui_pt_all_counties_IN_2017.csv', header = TRUE, skip = 1)
nui_2018 <- read.csv('Data/nui_pt_all_counties_IN_2018.csv', header = TRUE, skip = 1)
nui_2019 <- read.csv('Data/nui_pt_all_counties_IN_2019.csv', header = TRUE, skip = 1)
nui_2020 <- read.csv('Data/nui_pt_all_counties_IN_2020.csv', header = TRUE, skip = 1)

nui_2014_tb <- as.tibble(nui_2014)
nui_2015_tb <- as.tibble(nui_2015)
nui_2016_tb <- as.tibble(nui_2016)
nui_2017_tb <- as.tibble(nui_2017)
nui_2018_tb <- as.tibble(nui_2018)
nui_2019_tb <- as.tibble(nui_2019)
nui_2020_tb <- as.tibble(nui_2020)
```

I am next going to make sure the data is properly sorted:
```{r}
arrange(nui_2014_tb, X0)
arrange(nui_2015_tb, X0)
arrange(nui_2016_tb, X0)
arrange(nui_2017_tb, X0)
arrange(nui_2018_tb, X0)
arrange(nui_2019_tb, X0)
arrange(nui_2020_tb, X0)
```

Now I will join all the data together into one dataset:
```{r}
all_datasets <- left_join(nui_2014_tb, nui_2015_tb, by = c("X0", "NAME", "state", "county"))
all_datasets <- left_join(all_datasets, nui_2016_tb, by = c("X0", "NAME", "state", "county"))
all_datasets <- left_join(all_datasets, nui_2017_tb, by = c("X0", "NAME", "state", "county"))
all_datasets <- left_join(all_datasets, nui_2018_tb, by = c("X0", "NAME", "state", "county"))
all_datasets <- left_join(all_datasets, nui_2019_tb, by = c("X0", "NAME", "state", "county"))
all_datasets <- left_join(all_datasets, nui_2020_tb, by = c("X0", "NAME", "state", "county"))

all_datasets
```

Next we will perform some cleanup with the extra columns we have
```{r}
all_datasets_clean <- pivot_longer(
  data = all_datasets,
  cols = c(NUI_PT, NUI_PT.x, NUI_PT.y, NUI_PT.x.x, NUI_PT.y.y, NUI_PT.x.x.x, NUI_PT.y.y.y),
  names_to = "NUI_PT_Names",
  values_to = "NUI",
  values_drop_na = TRUE
)
all_datasets_clean <- pivot_longer(
  data = all_datasets_clean,
  cols = c(time, time.x, time.y, time.x.x, time.y.y, time.x.x.x, time.y.y.y),
  names_to = "time_Names",
  values_to = "year",
  values_drop_na = TRUE
)

```

Next, I will graph the number of uninsured people from 2014 to 2020 and try to find a trend.

From the below graph I am not seeing any trend amongst the years from 2014 to 2020.
```{r}
ggplot(data = all_datasets_clean, aes(x = year, y = NUI, color = county)) +
  geom_point() +
  xlab("Years") +
  ylab("Number of Uninsured from years 2014 to 2020")

```