---
title: "DSCI610_HW6_Cody_York"
output: html_notebook
---

```{r}
library(tidyverse)
library(readr)
library(VGAM)

df_child19 <- read_csv("Week 10/child19.csv")

df_child19_select_fields <- select(df_child19, INCGRP_C, RACEALLP_C, SDQPEER_C, SDQHYPE_C, SDQCOND_C, YRSINUS_C,
                                   PCNTADTWKP_C, SCFAMSTR_C, PARSEX1_C, RACPARSC_C, PARAGETC1_C, PARAGETC2_C, MARSTAT1_C,
                                   MARSTAT2_C, JAILEV_C, MENTDEPEV_C, FDSNOTEAT_C, BEHDFCNTR_C, ASDEV_C, OVER65FLG_C)
df_child19_select_fields_cleaned <- df_child19_select_fields %>% filter(!is.na(ASDEV_C) &
                                                                          !is.na(BEHDFCNTR_C) &
                                                                          !is.na(MENTDEPEV_C) &
                                                                          !is.na(JAILEV_C))
View(df_child19_select_fields)

```

# 1
The two binary responses I will be looking at are "Difficulty Controlling Behavior" and "Whether they have ever had autism".

For this first model I will check if having a parent in jail has an effect on "Difficulty Controlling Behavior".
For the second model I will check if having a parent with a mental illness has an effect on "Difficulty Controlling Behavior".

For this first model I will check if having a parent in jail has an effect on "a child having autism".
For the second model I will check if having a parent with a mental illness has an effect on "a child having autism".
```{r}
lm_beh_jail <- glm(JAILEV_C ~ BEHDFCNTR_C, family = binomial, data = df_child19_select_fields)
summary(lm_beh_jail)

lm_beh_ment <- glm(MENTDEPEV_C ~ BEHDFCNTR_C, family = binomial, data = df_child19_select_fields)
summary(lm_beh_ment)

lm_aut_jail <- glm(JAILEV_C ~ ASDEV_C, family = binomial, data = df_child19_select_fields)
summary(lm_aut_jail)

lm_aut_ment <- glm(MENTDEPEV_C ~ ASDEV_C, family = binomial, data = df_child19_select_fields)
summary(lm_aut_ment)

```

# 2.

The baseline category I will be looking at is "Income Group" .

```{r}
hometab <- ftable(xtabs(~INCGRP_C + RACEALLP_C + MARSTAT1_C + MARSTAT2_C, data = df_child19_select_fields))
ht <- as.data.frame(hometab)

w <- pivot_wider(ht, names_from = INCGRP_C, values_from = Freq)
#w
col_order <- c("RACEALLP_C", "MARSTAT1_C", "MARSTAT2_C", "Group1", "Group3", "Group5")
w2 <- w[, col_order]
w2

fit1 <- vglm(formula = cbind(Group1, Group3, Group5) ~ factor(RACEALLP_C) +
  factor(MARSTAT1_C) +
  factor(MARSTAT2_C),
             family = multinomial, data = w2)
summary(fit1)
```

# 3

This proportional odds model is for  "Difficulty Controlling Behavior".

```{r}
dptab <- ftable(xtabs(~INCGRP_C + RACEALLP_C + BEHDFCNTR_C, data = df_child19_select_fields))
dt1 <- as.data.frame(dptab)
#dt1
w1 <- pivot_wider(dt1, names_from = BEHDFCNTR_C, values_from = Freq)
#w1
col_order <- c("IncomeGroup", "Race", "Most", "Several", "None")
w3 <- w1[, col_order]
w3

fit2 <- vglm(formula = cbind(Most, Several, None) ~ factor(INCGRP_C) + factor(RACEALLP_C),
             family = cumulative(parallel = TRUE), data = w3)
summary(fit2)
```
