---
title: "DSCI_610_HW_1_Cody_York"
output: html_notebook
---

1. (2 pts) Generate two character vectors that hold the names of five countries from each continent and
their capitals respectively.
```{r}
countries <- c("United States", "India", "China", "Ireland", "Canada")

capitols <- c("Washington D.C.", "New Delhi", "Beijing", "Dublin", "Ottawa")

```

2. (5 pts) Generate a matrix with integers that has 12 rows and 5 columns. Treat the rows as 12 subjects.
Create a character vector that holds identification of these subjects. Feel free to combine letters and
numbers to create the identifications. Then include this vector and turn the matrix that you created
earlier into a data frame. Be creative in representing the columns with variables and assign suitable
column names.

```{r}
subject_grades <- matrix(40:100, nrow = 12, ncol = 5, byrow = FALSE)

subjects <- c("Math Grades", "English Grades", "Biology Grades", "Statistics Grades", "Geology Grades", "Art Grades",
              "P.E. Grades", "Reading Grades", "Music Grades", "Programming Grades", "Acting Grades",
              "Game Development Grades")

grades_df <- data.frame(subject_grades, subjects)

```
3. (5 pts) Write a function that takes a numeric vector as an argument and returns the standard deviation
of the elements in the vector.
```{r}

my_standard_deviation <- function(my_vector) {
  sd(my_vector)
}

my_standard_deviation(c(3, 10, 12, 12, 16, 19))
```

4. (8 pts) Consider the ‘PIdiabetes.csv‘ file. Apply a suitable loop function to find the mean and standard
deviation of first eight columns in the data set.
```{r}
PIdiabetes <- read.csv("PI_diabetes.csv", header = TRUE)

PIdiabetes_first_8_col <- PIdiabetes[, 1:8]

means_per_column <- lapply(PIdiabetes_first_8_col, mean)

sd_per_column <- lapply(PIdiabetes_first_8_col, sd)
```
